---
title: "wrangling-and-cleaning"
format: html
editor: visual
---

## load libraries

library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(dplyr)
library(tidyr)
library(readr)

## load data 

bit.ly/data-wrangle-lesson-data

catch_original <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", method = "libcurl"))

## data cleaning and wrangling 


## data expiration 

head(catch_original)
summary(catch_original)
unique(catch_original$Region)

```{r}
head(catch_original)
summary(catch_original)
unique(catch_original$Region)
# view(catch_original)
```

## Pipe operator 

remove "all" and "notersRedcode" columns 

```{r}
catch_data <- catch_original %>%
    select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)

head(catch_data)
```
```{r}
catch_data <- catch_original %>%
    select(-All,-notesRegCode)
```

```{r}
dplyr::glimpse(catch_data)
```

## change column content "mutate" from a string into a number
```{r}
catch_clean <- catch_data %>%
    mutate(Chinook = as.numeric(Chinook))
```
## 'is.na" will be false if ... and will show true when R finds an "NA"
```{r}
which(is.na(catch_clean$Chinook))
```

## pipe with mutate 
```{r}
catch_clean <- catch_data %>%
    mutate(Chinook = as.numeric(Chinook))
```

head(catch_clean)
i <- which(is.na(catch_clean$Chinook))
catch_data[i,]

# if else condition, if its an "i" write a "1"

```{r}
catch_clean <- catch_data %>%
    mutate(Chinook = if_else(condition = Chinook == "I", 
                             true = "1", 
                             false = Chinook),
           Chinook = as.numeric(Chinook))
```

##check if the "i" was removed 
catch_clean[i, ]

## pivot longer and pivot wider changes the view and presentation of the data      
## the negative "-c" says everything but/except 
```{r}
catch_long <- catch_clean %>% 
    #pivot longer all columns except Region and Year
    pivot_longer(cols = -c(Region, Year),
        names_to = "species",
        values_to = "catch")
```
head(catch_long)
```{r}
catch_wide <- catch_long %>%
    pivot_wider(names_from = species,
                values_from = catch)
```
head(catch_wide)

# renaming columns (the catch column to be called catch 1000s)
```{r}
catch_long <- catch_long %>%
    rename(catch_thousands = catch)
head(catch_long)

```


# adding columns with mutate

```{r}
catch_long <- catch_long %>%
  mutate(catch = catch_thousands * 1000)
head(catch_long)

```

## remove catch_thousands column 
```{r}
   catch_long <- catch_long %>% select(-catch_thousands)
head(catch_long)
```

 
## summary statistics: getting average catch per region 
## first group by region, then calculate the mean

bit.ly/3ppGUsQ

```{r}
mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch))

head(mean_region)
```

## filter: use long df and select rows from SSE region
```{r}
sse_catch<- catch_long %>%
    group_by(Region == "SSE") %>%
    summarize(n = n())
```

## excercise: 
1. "filter" to catches over one millon fish 
head(catch_long)
```{r}
m_catch<- catch_long %>%
    filter(catch > 1000000)

head(m_catch)
```

2. select rows from the SSE region that have information about Chinook Salmon

```{r}
sse_Chinook_catch<- catch_long %>%
    filter(Region == "SSE" & species == "Chinook")
head(sse_Chinook_catch)
```
## Sort data: recaluate mean catch by region and then sort by mean catch

```{r}
mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch)) %>%
    arrange(mean_catch)

head(mean_region)
```
## Splitting a column using separate() and unite()


                            
```{r}
sites_df <- data.frame(site = c("HAW-101",
                            "HAW-103",
                            "OAH-320",
                            "OAH-219",
                            "MAI-039"))
sites_df <- sites_df %>%
    separate(site, c("island", "site_number"), "-")
```
```{r}
sites_df %>%
    unite(code, island, site_number, sep="-")
```

## putting it all together 
```{r}
catch_original <- read_csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", method = "libcurl"))

mean_region <- catch_original %>%
  select(-All, -notesRegCode) %>% 
  mutate(Chinook = ifelse(Chinook == "I", 1, Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook)) %>% 
  pivot_longer(-c(Region, Year), 
               names_to = "species", 
               values_to = "catch") %>%
  mutate(catch = catch*1000) %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))

head(mean_region)
```

